### 2014-08-23 text/Chapter05.tex
Examples of structural comonads (Chapter 5)

\begin{example}[Structural liveness]
The structural coeffect algebra for liveness is formed by
$(L, \sqcap, \sqcup, \ident{L}, \ident{D}, \sqsubseteq)$, where $L=\{ \ident{L}, \ident{D} \}$ is 
the same two-point lattice as in the flat version, that is $\ident{D} \sqsubseteq \ident{L}$
with a join $\sqcup$ and meet $\sqcap$. 
\end{example}

@@

\subsection{Examples}
~
\newpage
~

\newcommand{\ctx}{\textit{ctx}}

\begin{figure*}[t]

@@

  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}}, \aclrd{\textbf{q}}}~\ident{lift}_{\cclrd{s}, \cclrd{s'}}~\ctx)

@@

        \atimes \alift{\cclrd{t},\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~\ctx = & (\emph{exch})\\
  \sem{\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~
  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{t},\cclrd{s}}, \aclrd{\textbf{q}}}~\ident{swap}_{\cclrd{t},\cclrd{s}}~\ctx)

@@

        \atimes \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}~\ctx = & (\emph{contr})\\
  \sem{\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~
  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}\,\cpar\,\cclrd{t}}, \aclrd{\textbf{q}}}~\ident{dup}_{\cclrd{s}, \cclrd{t}}~\ctx)

@@

\caption{Categorical semantics of the structural coeffect calculus}
\label{fig:struct-semantics}
\end{figure*}

~
\newpage

- variable access gets single-variable context and so it just projects the value
- application does not duplicate before splitting - it splits directly (to get two sub-contexts)
- abs is pretty much the same
- weakening just drops a value with ``ign'' annotation - we write this using pseuod-pattern matching, but it could be special operation in the category
- sub, exchange and contraction are quite similar - they extract and then merge
~ 
\newpage

@@

% alculus names and rule names
\newcommand{\clflt}{$\lambda_{\textnormal{\sffamily Cf}}$}
\newcommand{\clstr}{$\lambda_{\textnormal{\sffamily Cs}}$}

\newcommand{\ruleft}[1]{\textsf{\clflt\small{-T#1}}}
\newcommand{\rulefs}[1]{\textsf{\clflt\small{-S#1}}}
\newcommand{\rulest}[1]{\textsf{\clstr\small{-T#1}}}
\newcommand{\ruless}[1]{\textsf{\clstr\small{-S#1}}}
\newcommand{\rulessc}[1]{\textsf{\clstr\small{-SC#1}}}

\newcommand{\cobind}[1]{{#1}^\dagger}
\newcommand{\obj}[0]{obj}
\newcommand{\ccat}[0]{C}

\begin{figure*}[t]
\newcommand{\uu}[1]{\overline{#1}}
\begin{equation*}
\sem{\ctyp{r_1 \csprd \ldots \csprd r_n}{(x_1 : \tau_1 \times \ldots \times x_n : \tau_n)} \vdash e : \tau} : 
  \ctyp{r_1 \csprd \ldots \csprd r_n}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau
\end{equation*}
\begin{equation*}
\begin{array}{rcll}
\sem{\ctyp{0}{\Gamma} \vdash x_i : \tau_i} &=&
    \epsilon_0                                                                   &\quad$\ruless{var}$
\\[0.32em]
\sem{\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2} &=& 
    \Lambda (\sem{\ctyp{r \csprd s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2} \circ \cmerge_{r, s})    &\quad$\ruless{fun}$
\\[0.32em]
\sem{\ctyp{s \csprd (r \csvop t)}{(\Gamma_1, \Gamma_2)} \vdash e_1~e_2 : \tau} &=& 
    (\lambda (\gamma_1, \gamma_2) \rightarrow
    \sem{\ctyp{s}{\Gamma_1} \vdash e_1 : \ctyp{r}{\tau_1} \rightarrow \tau_2}~\gamma_1~
    (\cobind{\sem{\ctyp{t}{\Gamma_2}\vdash e_2 : \tau_1}}{t,r}~\gamma_2)) \circ \csplit_{s,(r \vee t)}  &\quad$\ruless{app}$
\\[0.32em]
\sem{\ctyp{r'}{\Gamma'} \vdash e : \tau} &=&
    \sem{\ctyp{r}{\Gamma} \vdash e : \tau} \circ \sem{\ctxtrans{\ctyp{r'}{\Gamma'}}{\ctyp{r}{\Gamma}}}       &\quad$\ruless{ctx}$
\end{array}
\end{equation*}

\begin{equation*}
\begin{array}{rclll}
\sem{\ctxtrans{\ctyp{r'\csprd s}{(\Gamma_1', \Gamma_2)}}{\ctyp{r \csprd s}{(\Gamma_1, \Gamma_2)}}}
&=& \cmerge_{r, s} \circ (\sem{\ctxtrans{\ctyp{r'}{\Gamma_1'}}{\ctyp{r}{\Gamma_1}}} \times \ident{id})~\circ \csplit_{r', s}\quad&\quad\quad\quad& % spaces to align
\textnormal{\rulessc{nest}}
\\[0.8em]
\sem{\ctxtrans{\ctyp{r\csprd s}{(\Gamma_1, \Gamma_2)} }{ \ctyp{s\csprd r}{(\Gamma_2, \Gamma_1)}}}
&=& \cmerge_{s, r} \circ \ident{swap} \circ \csplit_{r, s}&&
\textnormal{\rulessc{exch}}
\\[0.8em]
\sem{\ctxtrans{\ctyp{r \csprd 1}{(\Gamma, ())}}{\ctyp{r}{\Gamma}}}
&=&\ident{fst} \circ \csplit_{r, 1}&&
\textnormal{\rulessc{empty}}
\\[0.8em]
\sem{\ctxtrans{ \ctyp{r\csprd 0}{(\Gamma, x:\tau)} }{ \ctyp{r}{\Gamma}}}
&=&\ident{fst} \circ \csplit_{r, 0}&&
\textnormal{\rulessc{weak}}
\\[0.8em]
\sem{\ctxtrans{\ctyp{r \csvop s}{(x:\tau)}}{
    \ctyp{r \csprd s}{(x:\tau, x:\tau)}}}
&=&\cmerge_{r, s} \circ \cdupl_{r, s}&&
\textnormal{\rulessc{contr}}
\\[0.8em]
\sem{\ctxtrans{\ctyp{r\csprd (s\csprd t)}{(\Gamma_1, (\Gamma_2, \Gamma_3))}}{
    \ctyp{(r\csprd s)\csprd t}{((\Gamma_1, \Gamma_2), \Gamma_3)}}}\;
&=& \cmerge_{r \csprd s, t} \circ (\cmerge_{r, s} \times \ident{id}) \circ \ident{assoc}_1 \circ
        (\ident{id} \times \csplit_{s, t}) \circ \csplit_{r, s \csprd t}&&
\textnormal{\rulessc{assoc}}
\\[0.8em]
\sem{\ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{s}{\Gamma}}}
&=&
\iota_{r, s}
&&
\textnormal{\rulessc{sub}}
\end{array}
\end{equation*}

\caption{Categorical semantics for~\clstr}
\label{fig:structural-semantics}
\vspace{-0.5em}
\end{figure*}

\section{Semantics of structural coeffects}
\label{sec:structural-semantics}

The semantics of structural coeffect calculus \clstr~can be defined similarly to the semantics of
\clflt. The most notable difference is that the structure of coeffect tag now mirrors the structure
of the variable context. Thus an expression $\ctyp{r \csprd s}{(\Gamma_1, \Gamma_2)} \vdash e : \tau$ 
is modelled as a function $\ctyp{r \csprd s}{(\Gamma_1 \ctimes \Gamma_2)} \rightarrow \tau$.

As discussed in \ref{sec:structural-types}, the variable context $\Gamma$ in structural coeffect
system is not a simple finite product, but instead a binary tree. To model this, we do not use
ordinary products in the domain of the semantic function, but instead use a special 
constructor $\ctimes$. This way, we can guarantee that the variable structure corresponds to 
the tag structure.

\subsection{Structural tagged comonads}
\label{sec:structural-tagged-comonads}

To model composition of functions, we reuse the definition of \emph{tagged comonads} from 
Section~\ref{sec:tagged-comonads} without any change. This means that composing morphisms
$T^r \tau_1 \rightarrow \tau_2$ with $T^s \tau_2 \rightarrow \tau_3$ still gives us a morphism
$T^{r \csvop s} \tau_1 \rightarrow \tau_3$ and we use the $\csvop$ operation to combine the 
context-requirements.

However, functions that do not exist in context have only a single input variable (with a 
single corresponding tag). To model complex variable contexts, we need two additional
operations that allow manipulation with the variable context. Similarly to the model of \clflt,
we also require operations that model duplication and sub-coeffecting:

\begin{definition}[Structural tagged comonad]
Given a structural coeffect tag structure $(S, \csprd, \csvop, 0, 1)$ a \emph{structural tagged comonad} is a 
tagged comonad over $(S, \csvop, 0)$ comprising of $T^r$, $\epsilon_0$ and
$\cobind{(-)}{r, s}$ together with a mapping $- \ctimes -$ from a pair of objects 
$\obj{\ccat} \times \obj{\ccat}$ to an object $\obj{\ccat}$ and families of mappings:
%
\begin{equation*}
\begin{array}{rcl}
\cmerge_{r, s} &:& T^r A \times T^s B \rightarrow T^{(r \csprd s)}(A \ctimes B)\\
\csplit_{r, s} &:& T^{(r \csprd s)}(A \ctimes B) \rightarrow T^r A \times T^s N\\
\end{array}
\end{equation*}
%
% Delta is still defineable
% ?? \cdupl_{r, s}  &:& T^{(r \csvop s)}A \rightarrow T^r A \times T^s A\\
%
And with a family of mappings $\iota_{r, s} : T^r A \rightarrow T^s A$ for all $r, s \in S$ such that $r \vee s = r$.
%
%  TODO: laws?
%
\end{definition}

>>>>

\begin{example}[Structural liveness]
The structural coeffect algebra for liveness is formed by
$(\mathcal{L}, \sqcap, \sqcup, \ident{L}, \ident{D}, \sqsubseteq)$, where $\mathcal{L}=\{ \ident{L}, \ident{D} \}$ is 
the same two-point lattice as in the flat version, that is $\ident{D} \sqsubseteq \ident{L}$
with a join $\sqcup$ and meet $\sqcap$. 
\end{example}

@@

\begin{figure*}[!t]

@@

  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}}, \alift{\cclrd{s'}}, \aclrd{\textbf{q}}}~\ident{lift}_{\cclrd{s}, \cclrd{s'}}~\ctx)

@@

        \atimes \alift{\cclrd{t},\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~\ctx = & (\emph{exch})\\[-0.25em]
  \hspace{2em}  \sem{\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}
     {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\\[-0.25em]
  \hspace{4em}  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{t},\cclrd{s}}, 
     \alift{\cclrd{s},\cclrd{t}}, \aclrd{\textbf{q}}}~\ident{swap}_{\cclrd{t},\cclrd{s}}~\ctx)

@@

        \atimes \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}~\ctx = & (\emph{contr})\\[-0.25em]
  \hspace{2em}  \sem{\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}
      {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\\[-0.25em]
  \hspace{4em}  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}\,\cpar\,\cclrd{t}}, 
      \alift{\cclrd{s},\cclrd{t}}, \aclrd{\textbf{q}}}~\ident{dup}_{\cclrd{s}, \cclrd{t}}~\ctx)

@@

\caption{Categorical semantics of the structural coeffect calculus}
\label{fig:struct-semantics}
\end{figure*}

The concrete semantics for liveness and bounded variable use shown in 
Sections~\ref{sec:applications-struct-live} and \ref{sec:applications-struct-bll}
suggests that semantics of structural coeffect calculi tend to be more complex 
than semantics of flat coeffect calculi. The complexity comes from the fact
that we need a more expressive representation of the variable context -- \eg~a vector
of optional values -- and that the structural system needs to pass separate variable contexts
to the sub-expressions.

The latter aspect is fully captured by the semantics shown in this section. The
earlier point is left to the concrete notion of structural coeffect. Our model still gives
us the flexibility of defining the concrete representation of variable vectors. We explore
a number of examples in Section~\ref{sec:struct-semantics-examples} and start by looking
at the unified categorical semantics defined in terms of \emph{structural indexed comonads}.

\paragraph{Contexts and functions.}
In the structural coeffect calculus, expressions in context are interpreted as functions taking
a vector (represented using $-\xtimes-$) wrapped in a structure indexed with a vector of annotations
such as $\ctyp{\aclrd{\textbf{r}}}$. Functions take only a single variable as an input and so the
structure is annotated with a scalar, such as $\ctyp{\cclrd{r}}$, which we treat as being equivalent
to a singleton vector annotation $\ctyp{\alift{\cclrd{r}}}$:
%
\begin{equation*}
\begin{array}{rcl}
\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} } \vdash e : \tau} 
  &:& \ctyp{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} }{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau\\
\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2} &=& \ctyp{\alift{\cclrd{r}}}{\tau_1} \rightarrow \tau_2
\end{array}
\end{equation*}
%
Note that the instances of flat indexed comonad ignored the fact that the variable 
context wrapped in the data structure is a product. This is not generally the case for the 
structural indexed comonads -- the definitions shown in Section~\ref{sec:struct-semantics-examples}
are given specifically for $\ctyp{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} }{(\tau_1 \xtimes \ldots \xtimes \tau_n)}$
rather than more generally for $\ctyp{\aclrd{\textbf{r}}}{\alpha}$. The need for examining the structure
of the variable context is another reason for using $-\xtimes-$ when interpreting expressions in contexts.

\paragraph{Expressions.}
The semantics of strucutural coeffect calculi is shown in Figure~\ref{fig:struct-semantics}. As in
the previous chapter, the semantics is written in a programming language style using constructs
such as let-binding rather than using a categorical (point-free) notation. As before, the semantics
can be written using standard primitives (currying, uncurrying, function pairing etc.).

The following summarizes how the standard syntax-driven rules work, highlighting the differences
from the flat version:

\begin{itemize}
\item When accessing a variable (\emph{var}), the context now contains \emph{only} the accessed
  variable and so the semantics is just $\ident{counit}_{\cunit}$ without a projection.
  Constants (\emph{const}) are interpreted using a global dictionary $\delta$ as earlier.

\item The semantics of flat function application first duplicated the context so that the same 
  variables can be passed to both sub-expressions. This is no longer needed -- the 
  (\emph{app}) rule splits the variables \emph{including} the additional context into two parts.
  Passing the first context to the semantics of $e_1$ gives us a function
  $\ctyp{\alift{\cclrd{t}}}{\tau_1} \rightarrow \tau_2$.

  The argument for the function is obtained by applying $\ident{cobind}_{\cclrd{t}, \aclrd{\textbf{s}}}$ 
  to the semantics of $e_2$. The resulting function 
  $\ctyp{\cclrd{t}\aseq\aclrd{\textbf{s}}}{(\ldots \xtimes \ldots \xtimes \ldots)} \rightarrow \ctyp{\alift{\cclrd{t}}}{\tau_1}$
  is then called with the latter part of the context to obtain argument for the first function.

\item The semantic of function abstraction (\emph{abs}) is syntactically the same as in the 
  flat version -- the only difference is that we now merge a free-variable context with a 
  singleton vector, both at the level of variable assignments and at the level of coeffect 
  annotations.
\end{itemize}

The semantics for the non-syntax-driven rules performs transformations on the 
free-variable context. Weakening (\emph{weak}) splits the context and ignores the part corresponding
to the removed variable. If we were modelling the semantics in a language with a linear 
type system, this would require an additional operation for ignoring a context annotated with 
$\aclrd{\czero}$.

The remaining rules perform a transformation anywhere inside the free-variable vector. 
To simplify writing the semantics, we define a helper operation 
$\ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}$
that splits the variable vector into three parts, transforms the middle part and then 
merges them, using the newly transformed middle part.

The transformations on the middle part are quite simple. The (\emph{sub}) rule 
uses $\ident{lift}_{\cclrd{s}, \cclrd{s'}}$ to discard some of the available additional context;
the (\emph{exch}) rule swaps two single-variable contexts and the (\emph{contr}) rule
uses the $\ident{dup}_{\cclrd{s}, \cclrd{t}}$ operation to duplicate a varaible while 
splitting its additional context.

\paragraph{Properties.}

As in the flat calculus, the main reason for defining the categorical semantics in this chapter
is to provide validation for the design of the calculus. As we show in the next section, the 
discussed examples (liveness, data-flow, bounded variable reuse) form \emph{structural indexed
comonads} and so the calculus captures them correctly if the coeffect annotations in the typing
rules match the indices in the semantics. More formally:

\begin{remark}[Correspondence]
In all of the typing rules of the structural coeffect system, the context annotations $\aclrd{\textbf{r}}$ 
and $\cclrd{s}$ of typing judgements $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ and function types 
$\tau_1 \xrightarrow{\cclrd{s}} \tau_2$  correspond to the indices of mappings $\ctyp{\aclrd{\textbf{r}}}{}$ 
and $\ctyp{\alift{\cclrd{s}}}{}$ in the corresponding semantic function defined 
by $\sem{\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}$.
\end{remark}
\begin{proof}
By analysis of the semantic rules in Figure~\ref{fig:struct-semantics}.
\end{proof}

\noindent
As in the flat calculus, the primitive operations of the structural indexed comonad are all 
annotated with different operations provided by the coeffect annotations. This means that the
semantics uniquely determines the structure of the typing rules of the strucutral coeffect
calculus. Thanks to the correspondence between the product structure $\atimes$ of the annotations
and the variable context $\xtimes$, the correspondence property also guarantees that variable
values are split correctly, as required by the structural nature of the type system.

\subsection{Examples of structural indexed comonads}
\label{sec:struct-semantics-examples}

The categorical semantics for structural coeffect calculus can be easily instantiated to give
a semantics of a concrete calculus. In this section, we look at the three examples discussed
throughout this chapter -- structural liveness and data-flow and bounded variable reuse. 
Some aspects of the earlier two examples will be similar to flat versions discussed in
Section~\ref{sec:flat-semantics} -- they are based on the same data structures (option and a 
list, respectively), but the data structures are composed differently. Generally speaking -- 
rather than having a data structure over a product of variables, we now have 
a vector of variables over a specific data structure.

The abstract semantics does not specify how vectors of variables should be represented, so this can
vary in concrete instantiations. In all our examples, we represent a vector of variables as a 
product written using $\times$. To distinguish between products representing vectors and ordinary
products (\eg~a product of contexts returned by \ident{split}), we write vectors using 
$\langle a, \ldots, b \rangle$ rather than using parentheses.

\paragraph{Data-flow.}
It is interesting to note that the semantics of data-flow and bounded variable both keep
a product of multiple values for each variable, so they are both built around an \emph{indexed list}
data structure. However, their \ident{cobind} and \ident{dup} operations work differently. We start
by looking at the structure modelling data-flow computations (variables written in bold face such
as $\mathbf{a}_1$ range over vectors while $a_1$ ranges over individual values).

\begin{example}[Indexed list for data-flow]
The indexed list model of data-flow computations is defined over a structural coeffect
algebra $(\mathbb{N}, +, \mathit{max}, 0, 0, \leq)$. The data type $\ctyp{\langle n_1, \ldots, n_k\rangle}$
is indexed by required number of past variables for each individual variable. It is defined
over a vector of variables $\alpha_1 \xtimes \ldots \xtimes \alpha_k$ and it keeps a product
containing a current value followed by $n_i$ past values:
%
\begin{equation*}
\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \xtimes \ldots \xtimes \alpha_k)} = 
\underbrace{(\alpha_1 \times \ldots \times \alpha_1)}_{(n_1+1)-\textnormal{times}} \times \ldots \times
  \underbrace{(\alpha_k \times \ldots \times \alpha_k)}_{(n_k+1)-\textnormal{times}}
\end{equation*}
%
The mappings that define the structural indexed comonad include the \ident{split} and \ident{merge}
operations that are shared by the other two examples (discussed below):
%
\begin{equation*}
\begin{array}{l}
\ident{merge}_{\langle m_1, \ldots, m_k \rangle, \langle n_1, \ldots n_l \rangle} 
  (\langle \mathbf{a_1}, \ldots, \mathbf{a_k} \rangle, \langle \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle) = \\[-0.25em]
\quad \langle \mathbf{a_1}, \ldots, \mathbf{a_k}, \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle
\\[0.45em]
\ident{split}_{\langle m_1, \ldots, m_k \rangle, \langle n_1, \ldots n_l \rangle} 
   \langle \mathbf{a_1}, \ldots, \mathbf{a_k}, \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle = \\[-0.25em]
\quad (\langle \mathbf{a_1}, \ldots, \mathbf{a_k} \rangle, \langle \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle)
\end{array}
\end{equation*}
%
The remaining mappings that are required by structural indexed comonad and capture the
essence of data-flow computations are defined as:
%
\begin{equation*}
\begin{array}{l}
\ident{counit}_{0}~\langle \langle a_0 \rangle \rangle = a_0
\\[0.45em]
\ident{cobind}_{m, \langle n_1, \ldots, n_k \rangle}~f~
  \langle \langle a_{1,0}, \ldots a_{1,m+{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,m+{n_k}} \rangle \rangle = \\[-0.25em]
\quad \langle\langle\; f \langle \langle a_{1,0}, \ldots a_{1,{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,{n_k}} \rangle \rangle, ~\ldots~, \\[-0.25em]
\quad ~~~\; f \langle \langle a_{1,m}, \ldots a_{1,m+{n_1}} \rangle, \ldots, \langle a_{k,m}, \ldots a_{k,m+{n_k}} \rangle \rangle~\; \rangle\rangle
\\[0.45em]
\ident{dup}_{m, n} \langle \langle a_1, \ldots, a_{\mathit{max}(m,n)} \rangle \rangle =
   \langle \langle a_1, \ldots, a_m \rangle, \langle a_1, \ldots, a_n \rangle \rangle
\\[0.45em]
\ident{lift}_{k', k} \langle\langle a_0, \ldots, a_{k'} \rangle\rangle = 
  \langle\langle a_0, \ldots, a_k \rangle\rangle \qquad\qquad(\hspace{-0.95em}\textnormal{\footnotesize when}~k\leq k')
\end{array}
\end{equation*}
\end{example}

The definition of the indexed list data structure relies on the fact that the number of annotations
corresponds to the number of variables combined using $-\xtimes-$. It then creates a vector of lists
containing $n_i+1$ values for $i$-\textit{th} variable (the annotation represents the number of 
required \emph{past} values so one more value is required).

The \ident{split} and \ident{merge} operations are defined separately, because they are not specific
to the example. They operate on the top-level vectors of variables (without looking at the 
representation of the variable). This means that we can re-use the same definitions for the following
two examples (with the only difference that $\mathbf{a_i}, \mathbf{b_i}$ will represent options
rather than lists).

The mappings that explain how data-flow computations work are \ident{cobind} (representing sequential
composition) and \ident{dup} (representing context sharing or parallel composition). In \ident{cobind},
we get $k$ vectors corresponding to $k$ variables, each with $m+n_i$ values. The operation calls 
$f$ $m$-times to obtain $m$ past values required as the result of type $\ctyp{\langle m \rangle} \beta$.

The \ident{dup} operation needs to produce a two-varaible context containing $m$ and $n$ values,
respectively, of the input variable. The input provides $\mathit{max}(m, n)$ values, so the definition
is simply a matter of restriction. Finally, \ident{counit} extracts the (only) value of the (only) variable
and \ident{lift} drops additional past values that are not required.

\paragraph{Bounded reuse.}
As mentioned earlier, the semantics of calculus for bounded reuse is also based on the indexed list 
structure. Rather than representing possibly different past values that can be shared (\emph{c.f.}~\ident{dup}), 
the list now represents multiple copies of the same value that cannot be shared.

\begin{example}[Indexed list for bounded reuse]
The indexed list model of bounded variable reuse is defined over a structural coeffect
algebra $(\mathbb{N}, \ast, +, 1, 0, \leq)$. The data type $\ctyp{\langle n_1, \ldots, n_k\rangle}$ 
is a vector containing $n_i$ values of $i$-\textit{th} variable:
%
\begin{equation*}
\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \xtimes \ldots \xtimes \alpha_k)} = 
\underbrace{(\alpha_1 \times \ldots \times \alpha_1)}_{n_1-\textnormal{times}} \times \ldots \times
  \underbrace{(\alpha_k \times \ldots \times \alpha_k)}_{n_k-\textnormal{times}}
\end{equation*}
%
The \ident{merge} and \ident{split} operations are defined as in \emph{indexed list for data-flow}.
The operations that capture the behaviour of bounded reuse are defined as:
%
\begin{equation*}
\begin{array}{l}
\ident{counit}_{1}~\langle \langle a_0 \rangle \rangle = a_0
\\[0.45em]
\ident{lift}_{k', k} \langle\langle a_0, \ldots, a_{k'} \rangle\rangle = 
  \langle\langle a_0, \ldots, a_k \rangle\rangle \qquad\qquad(\hspace{-0.95em}\textnormal{\footnotesize when}~k\leq k')
\\[0.45em]
\ident{dup}_{m, n} \langle \langle a_1, \ldots, a_{m+n} \rangle \rangle = 
\quad \langle \langle a_1, \ldots, a_m \rangle, \langle a_{m+1}, \ldots, a_{m+n} \rangle \rangle
\\[0.45em]
\ident{cobind}_{m, \langle n_1, \ldots, n_k \rangle}~f~
  \langle \langle a_{1,0}, \ldots a_{1,m\ast{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,m\ast{n_k}} \rangle \rangle = \\[-0.25em]
\quad \langle\; f \langle \langle a_{1,0}, \ldots a_{1,{n_1 - 1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,{n_k - 1}} \rangle \rangle, ~\ldots~, \\[-0.25em]
\quad ~~ f \langle \langle a_{1,(m-1)\ast n_1}, \ldots a_{1,(m-1) \ast n_1} \rangle, \ldots, 
                   \langle a_{k,m\ast n_k - 1}, \ldots a_{k,m\ast n_k - 1} \rangle \rangle~\; \rangle
\end{array}
\end{equation*}
\end{example}

\noindent
The \ident{counit} and \ident{lift} operations are defined as previously -- variable access 
extracts the only value of the only variable and sub-coeffecting allows discarding multiple
copies of a value that are not needed.

In the bounded variable reuse system, variable sharing is annotated with $+$ (in contrast with
\textit{max} used in data-flow). The \indent{dup} operation thus splits the $m+n$ available
values between two vectors of length $m$ and $n$, without \emph{sharing} a value. The 
\ident{cobind} operation works similarly -- it splits $m\ast n_i$ available values of each
variable into $m$ vectors containing $n_i$ copies and then calls the $f$ function $m$-times to 
obtain $m$ resulting values without sharing any input value.

\paragraph{Liveness.}
In both data-flow and bounded reuse, the data type is defined as a vector of values obtained
by applying some parameterized data type (indexed list) to types of individual variables. We can
generalize this pattern and define $\ctyp{\langle l_1, \ldots, l_n \rangle}$ in terms of 
$D^{l}$ where $D^{l}$ is a simpler indexed data type. For liveness, the definition lets us 
reuse the mapping used when defining the semantics of flat liveness. However, we cannot fully
define the semantics of the structural version in terms of the flat version -- the 
\ident{cobind} operation is different and we need to provide the \ident{dup} operation.

\begin{example}[Structural indexed option]
Given a structural coeffect algebra formed by $(\{ \ident{L}, \ident{D} \}, \sqcap, \sqcup, \ident{L}, \ident{D}, \sqsubseteq)$
and the indexed option data type $D^l$, such that $D^{\ident{D}}\alpha = 1$ and $D^{\ident{L}}\alpha=\alpha$,
the data type for structural indexed option comonad is:
%
\begin{equation*}
\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \xtimes \ldots \xtimes \alpha_k)} = 
  D^{n_1} \alpha_1 \times \ldots \times D^{n_k} \alpha_k
\end{equation*}
%
The \ident{merge} and \ident{split} operations are defined as earlier. The remaining operations
model variable liveness as follows:
%
\begin{equation*}
\begin{array}{l}
\ident{cobind}_{\ident{L}, \langle l_1, \ldots, l_n \rangle}~f~\langle a_1, \ldots, a_n \rangle = \langle f~\langle a_1, \ldots, a_n \rangle \rangle \\[-0.25em]
\ident{cobind}_{\ident{D}, \langle \ident{D}, \ldots, \ident{D} \rangle}~f~\langle (), \ldots, () \rangle = \langle \ident{D} \rangle
\\[0.5em]
\hspace{-0.5em}
\begin{array}{lll}
~\\[-0.85em]
\ident{dup}_{\ident{D}, \ident{D}} \langle () \rangle &\narrow{=}& \langle (), () \rangle \\[-0.25em]
\ident{dup}_{\ident{L}, \ident{D}} \langle a \rangle &\narrow{=}& \langle a, () \rangle \\[-0.25em]
\ident{dup}_{\ident{D}, \ident{L}} \langle a \rangle &\narrow{=}& \langle (), a \rangle \\[-0.25em]
\ident{dup}_{\ident{L}, \ident{L}} \langle a \rangle &\narrow{=}& \langle a, a \rangle 
\end{array}
\qquad\quad
\begin{array}{lll}
\ident{counit}_{\ident{L}}~\langle a \rangle &\narrow{=}& a
\\[0.45em]
\ident{lift}_{\ident{L}, \ident{L}} \langle a \rangle &\narrow{=}& \langle a \rangle \\[-0.25em]
\ident{lift}_{\ident{L}, \ident{D}} \langle a \rangle &\narrow{=}& \langle () \rangle \\[-0.25em]
\ident{lift}_{\ident{D}, \ident{D}} \langle () \rangle &\narrow{=}& \langle () \rangle \\[-0.25em]
\end{array}
\end{array}
\end{equation*}
\end{example}

\noindent
When the expected result of the \ident{cobind} operation is dead (second case), the operation can
ignore all inputs and directly return the unit value $()$. Otherwise, it passes the vector of 
input variables to $f$ as-is -- no matter whether the individual values are live or dead. The
\ident{L} annotation is a unit with respect to $\cap$ and so the annotations expected by $f$ are
the same as those required by the result of \ident{cobind}.

The \ident{dup} operation resembles with the flat version of \ident{split} -- this is expected as
duplication in the flat calculus is performed by first duplicating the variable context (using 
\ident{map}) and then applying \ident{split}. Here, the duplication returns a pair which may or
may not contain value, depending on the annotations.

Finally, \ident{counit} extracts a value which is always present as guaranteed by the type
$\ctyp{\langle \ident{L} \rangle}{\alpha} \rightarrow \alpha$. The lifting operation models 
sub-coeffecting which may drop an available value (second case) or behaves as identity.

% ==================================================================================================
%                                                                   
%       ###                  #                    #       #          
%      #   #                 #                    #                  
%      #      #   #  # ##   ####    ###    ###   ####    ##     ###  
%       ###   #   #  ##  #   #         #  #   #   #       #    #   # 
%          #  #  ##  #   #   #      ####  #       #       #    #     
%      #   #   ## #  #   #   #  #  #   #  #   #   #  #    #    #   # 
%       ###       #  #   #    ##    ####   ###     ##    ###    ###  
%             #   #                                                  
%              ###                                                   
%
% ==================================================================================================

\section{Equational theory}
\label{sec:struct-syntax}

??

\newpage

@@

\newcommand{\clflt}{$\lambda_{\textnormal{\sffamily Cf}}$}
\newcommand{\clstr}{$\lambda_{\textnormal{\sffamily Cs}}$}
\newcommand{\rulest}[1]{\textsf{\clstr\small{-T#1}}}
